\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-20 1-intro.Rmd
further refine the motivation for a new spatio-temporal data structure

Spatio-temporal data concerns variables observed in the space across time and this paper focuses on the spatio-temporal vector data with fixed locations and regular time interval. This type of data can naturally been considered as a data cube and here a cube with three axes: spatial identifier, temporal identifier, and variable can be used to conceptualise spatio-temporal data. In this cubic formation, space only occupy one dimension, rather than two, to avoid the needs of hypercube for multivariate spatio-temporal data.

While conceptually the data format is relatively simple, the actual data object that comes to the analysts in the analysis can be in various forms. Figure \ref{fig:illu-input} shows four examples of wild spatio-temporal data: 1) combines the time invariant and varying variables into a single table through duplicating the time invariant variables; 2) splits those variables into a separate spatial table and a temporal table; 3) ...; and lastly 4) is array-based data, like NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

@@

Different authors have proposed various data objects to work with spatial temporal data in \proglang{R} and this includes: \pkg{spacetime} [@spacetime] is built on the \pkg{sp} [@sp] and \pkg{xts} [@xts] class and both of which have been replaced by more recent spatial and temporal data objects. \pkg{spatstat} [@spatstat] implements a `ppp` class for analysing point pattern data; and more recent, \pkg{stars} [@stars] has implemented an spatio-temporal array for wrangling both vector and raster data.

\textcolor{red}{Think again on the motivation and why it is hard to work existing tools.} In spatio-temporal data analysis, analysts will receive one of the various data formats depends on the data provider and need to manually rearrange the data to fit into one of the data object in \proglang{R} before starting any analysis on the data itself.

Looking at spatial or temporal data analysis, data objects like \pkg{sf} [@sf] and \pkg{tsibble} [@tsibble] have smoothed the data analysis in these two domains while there has not yet been a spatio-temporal data structure that makes it easy to explore spatio-temporal data and this creates a gap in the software development. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: climate observations from weather stations, air quality variables from air monitoring stations, or any variables that are measured at fixed location across time.

@@

This paper describes the implementation of a new spatio-temporal data structure: \pkg{cubble}. \pkg{Cubble} implements a data structure that uses two forms to manage the switch between spatial and temporal dimension. With this structure, users can manipulate the spatial or temporal dimension separately, while leaves the linking of two dimensions to \pkg{cubble}. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

>>>>

Spatio-temporal data concerns variables observed in the space across time and this paper focuses on the spatio-temporal vector data with fixed locations and regular time interval. This type of data can naturally been considered as a data cube and here a cube with three axes: spatial identifier, temporal identifier, and variable can be used to conceptualise spatio-temporal data. In this cubic formation, space only occupy one dimension, rather than two, to avoid the needs of hypercube for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the actual data object that comes to the analysts in the analysis can be in various forms. Figure \ref{fig:illu-input} shows four examples of wild spatio-temporal data: 1) combines the time invariant and varying variables into a single table through duplicating the time invariant variables; 2) splits those variables into a separate spatial table and a temporal table; 3) ...; and lastly 4) is array-based data, like NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

@@

Different authors have proposed various data objects to work with spatial temporal data in \proglang{R}. \pkg{spacetime} [@spacetime] is built on the \pkg{sp} [@sp] and \pkg{xts} [@xts] class and both of which have been replaced by more recent spatial and temporal data objects. \pkg{spatstat} [@spatstat] implements a `ppp` class for analysing point pattern data; and more recent, \pkg{stars} [@stars] has implemented an spatio-temporal array for wrangling both vector and raster data.

None of these three formats in Figure \ref{fig:illu-input} is not particularly convenient to work with. To augment a one-site-per-row summary with the original spatial variables in a single table format (example 1 in Figure \ref{fig:illu-input}), analysts will need to first make a distinct one-site-per-row summary of the original spatial variables before the join. This step adds few value to the analyais but is a tedious step that first select all the spatial variables and then find the distinct row of each. In another type of operation that filter the sites based on per-site summary, intermediate objects, like the per-station summary and the filtered station id vector, will need be created as standalone objects in the environment. The spirit of using pipe reduces the intermediate data objects that are unnecessarily created in the analysis while in this case, these only-used-once data objects are hard to be incorporated into a piping workflow. In addition, if the data is organised as a separate spatial and temporal table as example 2 in Figure \ref{fig:illu-input}, the filter on site ID will need to performed twice in each table to keep them both up-to-date. With the array data structure, although it can easily store and wrangle large spatio-temporal data, the 3D or higher array can be less intuitive than a 2D tabular data in data wrangling.

Data objects like \pkg{sf} [@sf] and \pkg{tsibble} [@tsibble] have smoothed the data analysis in spatial or temporal data analysis, while there has not yet been a spatio-temporal data structure that makes it easy to explore spatio-temporal data and this creates a gap in the software development. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: climate observations from weather stations, air quality variables from air monitoring stations, or any variables that are measured at fixed location across time.

@@

This paper describes the a new tabular data object:  \pkg{cubble} for wrangling spatio-temporal data. \pkg{Cubble} overcomes the inconvenience in data wrangling mentioned above through two forms: a nested form to organise spatial variables and a long form to organise temporal variables. With this structure, analysts can explore the spatio-temporal data with less efforts spend on organising variables at different observational units. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

### 2022-02-20 main.tex
further refine the motivation for a new spatio-temporal data structure

Spatio-temporal data concerns variables observed in the space across
time and this paper focuses on the spatio-temporal vector data with
fixed locations and regular time interval. This type of data can
naturally been considered as a data cube and here a cube with three
axes: spatial identifier, temporal identifier, and variable can be used
to conceptualise spatio-temporal data. In this cubic formation, space
only occupy one dimension, rather than two, to avoid the needs of
hypercube for multivariate spatio-temporal data.

While conceptually the data format is relatively simple, the actual data
object that comes to the analysts in the analysis can be in various
forms. Figure \ref{fig:illu-input} shows four examples of wild
spatio-temporal data: 1) combines the time invariant and varying
variables into a single table through duplicating the time invariant
variables; 2) splits those variables into a separate spatial table and a
temporal table; 3) \ldots; and lastly 4) is array-based data, like
NetCDF (Network Common Data Form) data, which is commonly used in earth
science, e.g.~climatology, meteorology, and oceanography, among others,
to organise multivariate data in a spatio-temporal grid.

@@

Different authors have proposed various data objects to work with
spatial temporal data in \proglang{R} and this includes: \pkg{spacetime}
\citep{spacetime} is built on the \pkg{sp} \citep{sp} and \pkg{xts}
\citep{xts} class and both of which have been replaced by more recent
spatial and temporal data objects. \pkg{spatstat} \citep{spatstat}
implements a \texttt{ppp} class for analysing point pattern data; and
more recent, \pkg{stars} \citep{stars} has implemented an
spatio-temporal array for wrangling both vector and raster data.

\textcolor{red}{Think again on the motivation and why it is hard to work existing tools.}
In spatio-temporal data analysis, analysts will receive one of the
various data formats depends on the data provider and need to manually
rearrange the data to fit into one of the data object in \proglang{R}
before starting any analysis on the data itself.

Looking at spatial or temporal data analysis, data objects like \pkg{sf}
\citep{sf} and \pkg{tsibble} \citep{tsibble} have smoothed the data
analysis in these two domains while there has not yet been a
spatio-temporal data structure that makes it easy to explore
spatio-temporal data and this creates a gap in the software development.
The requirement for such a tool is important given the ubiquity of
spatio-temporal vector data in the wild: climate observations from
weather stations, air quality variables from air monitoring stations, or
any variables that are measured at fixed location across time.

This paper describes the implementation of a new spatio-temporal data
structure: \pkg{cubble}. \pkg{Cubble} implements a data structure that
uses two forms to manage the switch between spatial and temporal
dimension. With this structure, users can manipulate the spatial or
temporal dimension separately, while leaves the linking of two
dimensions to \pkg{cubble}. The software is available from the
Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

>>>>

Spatio-temporal data concerns variables observed in the space across
time and this paper focuses on the spatio-temporal vector data with
fixed locations and regular time interval. This type of data can
naturally been considered as a data cube and here a cube with three
axes: spatial identifier, temporal identifier, and variable can be used
to conceptualise spatio-temporal data. In this cubic formation, space
only occupy one dimension, rather than two, to avoid the needs of
hypercube for multivariate spatio-temporal data. While conceptually the
data format is relatively simple, the actual data object that comes to
the analysts in the analysis can be in various forms. Figure
\ref{fig:illu-input} shows four examples of wild spatio-temporal data:
1) combines the time invariant and varying variables into a single table
through duplicating the time invariant variables; 2) splits those
variables into a separate spatial table and a temporal table; 3) \ldots;
and lastly 4) is array-based data, like NetCDF (Network Common Data
Form) data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

@@

Different authors have proposed various data objects to work with
spatial temporal data in \proglang{R}. \pkg{spacetime} \citep{spacetime}
is built on the \pkg{sp} \citep{sp} and \pkg{xts} \citep{xts} class and
both of which have been replaced by more recent spatial and temporal
data objects. \pkg{spatstat} \citep{spatstat} implements a \texttt{ppp}
class for analysing point pattern data; and more recent, \pkg{stars}
\citep{stars} has implemented an spatio-temporal array for wrangling
both vector and raster data.

None of these three formats in Figure \ref{fig:illu-input} is not
particularly convenient to work with. To augment a one-site-per-row
summary with the original spatial variables in a single table format
(example 1 in Figure \ref{fig:illu-input}), analysts will need to first
make a distinct one-site-per-row summary of the original spatial
variables before the join. This step adds few value to the analyais but
is a tedious step that first select all the spatial variables and then
find the distinct row of each. In another type of operation that filter
the sites based on per-site summary, intermediate objects, like the
per-station summary and the filtered station id vector, will need be
created as standalone objects in the environment. The spirit of using
pipe reduces the intermediate data objects that are unnecessarily
created in the analysis while in this case, these only-used-once data
objects are hard to be incorporated into a piping workflow. In addition,
if the data is organised as a separate spatial and temporal table as
example 2 in Figure \ref{fig:illu-input}, the filter on site ID will
need to performed twice in each table to keep them both up-to-date. With
the array data structure, although it can easily store and wrangle large
spatio-temporal data, the 3D or higher array can be less intuitive than
a 2D tabular data in data wrangling.

Data objects like \pkg{sf} \citep{sf} and \pkg{tsibble} \citep{tsibble}
have smoothed the data analysis in spatial or temporal data analysis,
while there has not yet been a spatio-temporal data structure that makes
it easy to explore spatio-temporal data and this creates a gap in the
software development. The requirement for such a tool is important given
the ubiquity of spatio-temporal vector data in the wild: climate
observations from weather stations, air quality variables from air
monitoring stations, or any variables that are measured at fixed
location across time.

This paper describes the a new tabular data object: \pkg{cubble} for
wrangling spatio-temporal data. \pkg{Cubble} overcomes the inconvenience
in data wrangling mentioned above through a nested form to organise
spatial variables and a long form to organise temporal variables. With
this structure, analysts can explore the spatio-temporal data with less
efforts spend on \ldots{} The software is available from the
Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.
